Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-13 paper/paper.tex
the rest of bovy comments

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\data$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{\frac{1}{5}\,J^{\corr}_n} = 10^{\frac{1}{5}\,M^{\corr}_{J,n} + 2}
\label{eq:transform}
\end{equation}
where, for this expression, it is important to remember that $\varpi$ is assumed to be in units of $[mas]$.
The transformation of the dust corrected absolute magnitude is required to keep the uncertainties Gaussian, and Gaussian uncertainties are required to use \xd, see below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\truth_n = [10^{0.2\,M^{\true}_{J,n} + 2}, (J-K_s)^{\true}_n] \\
\mathrm{data \, vector} \quad &\data_n = [\varpi_n\,10^{0.2J^{\corr}_n}, (J-K_s)^{\corr}_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^{\corr}_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J-K_s)^{\corr}_n = J_n - K_{s,n} - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

@@

To infer $\varpi_{\true}$ once the \cmd\ prior is constructed, we first do the inference in the 2D space of the \cmd\ prior $p(\truth \given \alpha)$, using Equation \ref{eq:posterior}. We then project the 2D posterior into the 1D transformed absolute magnitude space, and do a little algebra to solve for $\varpi_{\true}$

\begin{eqnarray}
p(\truth_0 \given \data, \hat{\alpha}) &\rightarrow& p(\varpi_{\true} \given y, \hat{\alpha}) \, f(\varpi_{\true}) \\
\mathrm{where} \; f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{\true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}
with $f(\varpi_{\true})$ being a window function to insure $\varpi_{\true}$ is positive and to put $\varpi_{\true}$ on a similar grid for all the posteriors.

@@

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \gaia\ catalog. This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
And $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, or have a fraction decrease in variance by a factor of $>4$, shown as the other dotted line at $y=0.14$.

@@

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{m67.pdf}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more clustered and sharply peaked around the ``true" distance to M67 than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

@@

We obtained the photometric parallaxes in this project
by building a data-driven model of \cmd\ of the stars in the \tgas\ data set,
and using it as a prior PDF for Bayesian inference.
The posterior PDFs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
can introduce bias).
We have shown that, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well:
The posterior distance estimates to stars in (at least one) stellar
cluster are much more clustered than likelihood-based distance
estimates.

@@

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{comparePrior.pdf}
\caption{Comparison with a common spatial prior: A visualization of the posterior PDF for a subsection of \tgas\ stars, shown as the expectation value and the sqrt of the variance as the uncertainty. Left panel shows the posterior PDF using our \cmd\ prior, right panel shows the posterior PDFs when using the exponentially declining spatial density, a common space-density prior used when inferring distances. The scale of the exponentially declining spatial density prior is $1.35~{\rm kpc}$, the optimal value found in \cite{astraatmadja16a}. Using the \cmd\ as a prior generates a posterior PDF that has smaller variance.}
\label{fig:comparePrior}
\end{figure}

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so that the posteriors are also very smooth.
It is a fairly weak prior.
It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax.
\figurename~\ref{fig:comparePrior} shows a comparison of the EDSD prior with our \cmd\ prior.
Here the parallax posterior PDF expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.

@@

The lower main sequence is very narrow while the upper main sequence
is much wider. This may be an age variation on the main sequence mixed
with a Malmquist bias of seeing a much larger volume of upper main
sequence stars \citep{malmquist22}. There is a plausible binary sequence of stars with
about twice the brightness as the main sequence. The helium burning
red clump is prominent for the volume DR1 probed, and the red giant
branch is fairly straight and narrow. The main sequence turn off is
noticeable and also surprisingly narrow, possibly a reflection of the
star formation history of the local volume also mixed with a Malmquist
bias.

@@

  \cmd\ space. Thus, this has a small effect on our results, and could be improved by modelling the noise distribution simultaneously with the \cmd.

@@

  \tgas\ Catalog (or really the \tgas--\tmass\ intersection), not of

>>>>

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\data$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{\frac{1}{5}\,J^{\corr}_n} = 10^{\frac{1}{5}\,M^{\corr}_{J,n} + 2}
\label{eq:transform}
\end{equation}
where, for this expression, it is important to remember that $\varpi$ is assumed to be in units of $[mas]$.
The transformation of the (dust corrected) absolute magnitude is required to keep the uncertainties Gaussian, and Gaussian uncertainties are required to use \xd, see below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\truth_n = [10^{0.2\,M^{\true}_{J,n} + 2}, (J-K_s)^{\true}_n] \\
\mathrm{data \, vector} \quad &\data_n = [\varpi_n\,10^{0.2J^{\corr}_n}, (J-K_s)^{\corr}_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^{\corr}_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J-K_s)^{\corr}_n = J_n - K_{s,n} - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

@@

To infer $\varpi_{\true}$ once the \cmd\ prior is constructed, we first do the inference in the 2D space of the \cmd\ prior $p(\truth \given \alpha)$, using Equation \ref{eq:posterior}. We then project the 2D posterior into the 1D $\varpi_{\true}$ space by multiply by the Jacobian to calculate the posterior PDF over $\varpi_{\true}$

\begin{eqnarray}
p(\varpi_{\true} \given \data, \hat{\alpha}) \, &\propto& \, \left|\frac{\dd\truth}{\dd\varpi}\right| \, p(\truth \given \data, \hat{\alpha}) \, f(\varpi_{\true}) \\
%p(\truth_0 \given \data, \hat{\alpha}) &\rightarrow& p(\varpi_{\true} \given y, \hat{\alpha}) \, f(\varpi_{\true}) \\
\mathrm{where} \; f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{\true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}
with $f(\varpi_{\true})$ being a window function to insure $\varpi_{\true}$ is positive and to put $\varpi_{\true}$ on a similar grid for all the posteriors.

@@

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \gaia\ catalog. This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
The precision of our posterior PDFs is greater than or equal to the precision of the \gaia\ catalogue for $87\%$ of the stars, shown as the dotted line at $x=0$,
and $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, which corresponds to a fraction decrease in variance by a factor of $>4$, shown as the dotted line at $y=0.14$.

@@

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{m67.pdf}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more narrow and clustered than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

@@

We obtained the photometric parallaxes in this project
by building a data-driven model of \cmd\ of the stars in the \tgas\ data set,
and using it as a prior PDF for Bayesian inference.
The posterior PDFs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
can introduce bias).
We have shown that, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well:
The posterior distance estimates to stars in (at least one) stellar
cluster are much more clustered than likelihood-based distance
estimates, and they cluster around a sensible value.
Accuracy is demonstrated by this test because each star in the cluster is
subject to a different, unique distance prior (see \figurename~\ref{fig:posterior})
and yet the posterior PDFs are all pulled to the same parallax value.

@@

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{comparePrior.pdf}
\caption{Comparison with a common spatial prior: A visualization of the posterior PDF for a subsection of \tgas\ stars, shown as the expectation value and the square root of the variance as the uncertainty. Left panel shows the posterior PDF using our \cmd\ prior, right panel shows the posterior PDFs when using the exponentially declining spatial density, a common space-density prior used when inferring distances. The scale of the exponentially declining spatial density prior is $1.35~{\rm kpc}$, the optimal value found in \cite{astraatmadja16a}. Using the \cmd\ as a prior generates a posterior PDF that has smaller variance.}
\label{fig:comparePrior}
\end{figure}

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so that the posteriors are also very smooth.
It is a fairly weak prior.
\figurename~\ref{fig:comparePrior} shows a comparison of the EDSD prior with our \cmd\ prior.
Here the parallax posterior PDF expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.

@@

The lower main sequence is very narrow while the upper main sequence
is much wider. This may be an age and metallicity variation on the main sequence mixed
with a Malmquist bias of seeing a much larger volume of upper main
sequence stars \citep{malmquist22}. There is a plausible binary sequence of stars with
about twice the brightness as the main sequence. The helium burning
red clump is prominent for the volume DR1 probed, and the red giant
branch is fairly straight and narrow. The main sequence turn off is
noticeable and also surprisingly narrow, possibly a reflection of the
star formation history of the local volume also mixed with a Malmquist
bias.

@@

  \cmd\ space. Thus, this has a small effect on our results, and could be improved by modeling the noise distribution simultaneously with the \cmd.

@@

  \tgas\ Catalog (or really the \tgas--\tmass--\psone\ intersection), not of
